% R Bootcamp, Module 8: Graphics
% August 2016, UC Berkeley
% Kellie Ottoboni (kellieotto@berkeley.edu)
% Built off of material by Chris Krogslund

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(plyr)
library(dplyr)

library(tidyr)
library(reshape2)
library(ggplot2)
library(lattice)
library(gridExtra)
if(!('modules' %in% unlist(strsplit(getwd(), split = '/')))) setwd('modules')
data <- read.csv("../data/cpds.csv")
medals <- read.csv("../data/all_medalists.csv", header = TRUE, stringsAsFactors = FALSE)

```

# By way of introduction...

* 3 main facilities for producing graphics in R: **base**, **lattice**, and **ggplot2**
* In practice, these facilities are grouped into two camps: "basic" and "advanced"
* A better formulation: quick/dirty v. involved/fancy

# The data

* All Summer Olympic medalists from 1896-2008
* Variables include name, gender, country, sporting event, and type of medal won
* We don't actually care about names of winners: we're interested in how many medals that different countries won

For more info, see [The Guardian Datablog Olympic medal winners: every one since 1896 as open data](https://www.theguardian.com/sport/datablog/2012/jun/25/olympic-medal-winner-list-data)

First, we'll use dplyr and tidyr to count the medals of each type won, for each country and each year.
```{r}
head(medals)

# dplyr and tidyr refresher
medal_counts <- medals %>% group_by(Medal, Year, Country) %>% summarise(count = n())
head(medal_counts)
```

This table is in tidy format. Wide (untidy) format can be useful for plotting in base plot (more on this later)
```{r}
medal_counts_wide <- medal_counts %>% spread(key = Medal, value = count) %>%
  ungroup() %>%
  mutate(Bronze = ifelse(is.na(Bronze), 0, Bronze)) %>% 
  mutate(Silver = ifelse(is.na(Silver), 0, Silver)) %>% 
  mutate(Gold = ifelse(is.na(Gold), 0, Gold)) 
head(medal_counts_wide)
```

Finally, let's subset the data to gold medal counts for the US, for easier plotting.
```{r}
usa_gold_medals <-  medal_counts %>% 
  filter(Medal == "Gold") %>%
  filter(Country == "United States")
```

# Base graphics

The general call for base plot looks something like this:
  
```{r, eval=FALSE}
plot(x=, y=, ...)
```
Additional parameters can be passed in to customize the plot:

* type: scatterplot? lines? etc
* main: a title
* xlab, ylab: x-axis and y-axis labels
* col: color, either a string with the color name or a vector of color names for each point

More layers can be added to the plot with additional calls to `lines`, `points`, `text`, etc.

```{r}
plot(medal_counts_wide$Year, medal_counts_wide$Gold) # Basic
plot(usa_gold_medals$Year, usa_gold_medals$count, type = "l", 
     main = "USA Gold Medals",
     xlab = "Year", ylab = "Count") # with updated parameters
points(x = 1984, y = usa_gold_medals$count[usa_gold_medals$Year == 1984], 
       col = "red", pch = 16)
```

## Other plot types

These are just a few other types of plots you can make in base graphics.

```{r}
boxplot(Gold~Year, data = medal_counts_wide)
hist(medal_counts_wide$Gold)
plot(density(medal_counts_wide$Gold))
barplot(usa_gold_medals$count, width = 4, names.arg = usa_gold_medals$Year, main = "USA Gold Medals")
mosaicplot(Year~Medal, medal_counts)

```

## Object-oriented plots
* Base graphics often recognizes the object type and will implement specific plot methods
* lattice and ggplot2 generally **don't** exhibit this sort of behavior

```{r, fig.cap=" "}
medal_lm <- lm(Gold ~ Bronze + Silver, data = medal_counts_wide)
plot(medal_counts_wide %>% select(-Country)) # Calls plotting method for class of the dataset ("data.frame")
plot(medal_lm, which=1:2) # Calls plotting method for class of medal_lm object ("lm"), print first two plots only


```


## Pros/cons
* Good for exploratory data analysis and sanity checks
* Syntax is inconsistent across functions: some take x,y while others take formulas
* Defaults plotting parameters are ugly and it can be difficult to customize

# lattice and ggplot2

lattice is 

a) faster (though only noticeable over many and large plots) 

b) simpler (at first)

c) better at trellis graphs

d) able to do 3d graphs

ggplot2 is 

a) generally more elegant

b) more syntactically logical (and therefore simpler, once you learn it)

c) better at grouping

d) able to interface with maps

# Basic usage: lattice

The general call for lattice graphics looks something like this:
  
```{r, eval=FALSE}
graph_type(formula, data=, [options])
```

The most common graph types:
* xyplot: generic scatterplot
* barchart
* bwplot: boxplot
* histogram
* cloud: 3D scatterplot

The specifics of the **formula** differ for each graph type, but the general format is straightforward

```{r, eval=FALSE}
y             # Show the distribution of y

y~x           # Show the relationship between x and y 

y~x|A         # Show the relationship between x and y conditional on the values of A

y~x|A*B       # Show the relationship between x and y conditional on the combinations of A and B

z~y*x         # Show the 3D relationship between x, y, and z
```

Let's recreate some of the plots we made above using lattice. The options for lattice overlap with the options for base graphics.
```{r}
xyplot(Gold~Year, data = medal_counts_wide, main = "Gold Medal Counts")
xyplot(count~Year, data = usa_gold_medals, type = "l", main = "USA Gold Medals")

bwplot(Year~Gold, data = medal_counts_wide, main = "Gold Medal Counts") # Labels are lost when lattice automatically converts Year to a factor
bwplot(Gold~Year, data = medal_counts_wide, main = "Gold Medal Counts", horizontal = FALSE) # Flip the coordinates so year increases on the x-axis

histogram(~Gold, medal_counts_wide, type="density", main = "Gold Medal Counts")
densityplot(~Gold, medal_counts_wide, main = "Gold Medal Counts")
barchart(count~Year, data = usa_gold_medals, width = 4, main = "USA Gold Medals", horizontal = FALSE)
```

## Pros/cons  
* lattice tends to focus on a particular type of graph and how to represent cross-sectional variation by splitting it up into smaller chunks
* Becoming a proficient user of lattice requires learning a huge array of graph-specific formulas and options


# Basic usage: ggplot2

The general call for ggplot2 graphics looks something like this:
  
```{r, eval=FALSE}
ggplot(data=, aes(x=,y=, [options])) + geom_xxxx() + ... + ... + ...
```

Note that ggplot2 graphs in layers in a *continuing call* (hence the endless +...+...+...), which really makes the extra layer part of the call

```{r, eval=FALSE}
...+geom_xxxx(data=, aes(x=,y=,[options]),[options])+...+...+...
```
You can see the layering effect by comparing the same graph with different colors for each layer

```{r, fig.cap=" ", warning=FALSE}
p <- ggplot(data=medal_counts_wide, aes(x=Year, y=Gold)) + geom_point(color="gold")
p
p + geom_point(aes(x=Year, y=Silver), color="gray") + ylab("Medals")
```

## Grammar of Graphics

ggplot2 syntax is very different from base graphics and lattice. It's built on the **grammar of graphics**.
The basic idea is that the visualization of all data requires four items:

1) One or more **statistics** conveying information about the data (identities, means, medians, etc.)

2) A **coordinate system** that differentiates between the intersections of statistics (at most two for ggplot, three for lattice)

3) **Geometries** that differentiate between off-coordinate variation in *kind*
  
4) **Scales** that differentiate between off-coordinate variation in *degree*
  
ggplot2 allows the user to manipulate all four of these items.


## Anatomy of aes()

```{r, eval=FALSE}
ggplot(data=, aes(x=, y=, color=, linetype=, shape=, size=))
```

ggplot2 is optimized for showing variation on all four aesthetic types

```{r, fig.cap=" ", warning=FALSE}
# Differences in kind using color
pigs.data <- data[data$country %in% c("Greece", "Portugal", "Ireland", "Spain"),]
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_line(aes(color=country))

usa_all_medals <- medal_counts %>% filter(Country == "United States")
ggplot(data = usa_all_medals, aes(x = Year, y = count)) + geom_line(aes(color = Medal))
```

Note what happens when we specify the color parameter outside of the aesthetic operator. ggplot2 views these specifications as invalid graphical parameters. 

```{r, fig.cap=" ", warning=FALSE, error = TRUE}
ggplot(data = usa_all_medals, aes(x = Year, y = count)) + geom_line(color = Medal)
ggplot(data = usa_all_medals, aes(x = Year, y = count)) + geom_line(color = "Medal")
ggplot(data = usa_all_medals, aes(x = Year, y = count)) + geom_line(color = "red")
```

## ggplot2 and tidy data

* Base graphics/lattice and ggplot2 have one big difference: ggplot2 **requires** your data to be in tidy format. For base graphics, it can actually be helpful *not* to have your data in tidy format. Compare:

```{r}
usa_all_medals_untidy <- medal_counts_wide %>% filter(Country == "United States") # Medal is the variable, but it spans multiple columns

# Base graphics call
plot(usa_all_medals_untidy$Year, usa_all_medals_untidy$Gold, col = "green", type = "l")
lines(usa_all_medals_untidy$Year, usa_all_medals_untidy$Silver, col = "blue")
lines(usa_all_medals_untidy$Year, usa_all_medals_untidy$Bronze, col = "red")
legend("right", legend = c("Gold", "Silver", "Bronze"), fill = c("green", "blue", "red"))

# ggplot2 call
ggplot(data = usa_all_medals, aes(x = Year, y = count)) + geom_line(aes(color = Medal))
```

* The difference is that ggplot treats `Medal` as an aesthetic parameter that differentiates kinds of statistics. Base graphics treats each medal type as a layer to the plot.
* ggplot2 plays nice with dplyr and pipes. If you want to manipulate your data specifically for one plot but not save the new dataset, you can call your dplyr chain and pipe it directly into a ggplot call:

```{r}
ggplot(usa_gold_medals, aes(x = Year, y = count)) + geom_line()

# This combines the subsetting and plotting into one step
medal_counts %>% 
  filter(Medal == "Gold") %>%
  filter(Country == "United States") %>%
  ggplot(aes(x = Year, y = count)) + geom_line()
```

# Comparing lattice and ggplot

* Density Plots
* Scatter Plots
* Line Plots
* Bar plots
* Box plots
* Trellis Plots
* Contour Plots
* Tile/Image Plots
* 3D Plots (lattice)
* Panel Plots (ggplot2)

# lattice v. ggplot2: Densities

```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
ggplot(data = usa_gold_medals, aes(x = count)) + geom_density() # ggplot2
densityplot(~count, data = usa_gold_medals) # lattice
```

# lattice v. ggplot2: X-Y scatter plots

```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
ggplot(data = usa_gold_medals, aes(x = Year, y = count)) + geom_point() # ggplot2
xyplot(count~Year, data = usa_gold_medals) # lattice
```

# lattice v. ggplot2: X-Y line plots

```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
ggplot(data = usa_gold_medals, aes(x = Year, y = count)) + geom_line() # ggplot2
xyplot(count~Year, data = usa_gold_medals, type = "l") # lattice
```

# lattice v. ggplot2: bar plots 
```{r, fig.cap=" ", warning=FALSE, message=FALSE, fig.show = "hold"}
# Create a dataframe of median number of gold medals by country
median_gold_medals <- medal_counts %>% 
  filter(Medal == "Gold") %>% 
  group_by(Country) %>%
  summarise(med = median(count))

ggplot(data = median_gold_medals[1:15, ], aes(x = Country, y = med)) + geom_bar(stat="identity") # ggplot2
barchart(med~Country, data = median_gold_medals[1:15, ]) # lattice
```

# lattice v. ggplot2: box plots 
```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
# Notice that here, you must explicitly convert numeric years to factors
ggplot(data = medal_counts_wide, aes(x = factor(Year), y = Gold)) + geom_boxplot() # ggplot2
bwplot(Gold~factor(Year), data = medal_counts_wide, horizontal = FALSE) # lattice
```

# lattice v. ggplot2: "trellis" plots 
```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
# Subset the data to North America countries for easier viewing
northern_hem <- medal_counts_wide %>% filter(Country %in% c("United States", "Canada", "Cuba", "Mexico"))

ggplot(data = northern_hem, aes(x = Year, y = Gold)) + geom_point() + facet_wrap(~Country) # ggplot2
xyplot(Gold~Year|Country, data = northern_hem) # lattice
```

# lattice v. ggplot2: countour plots
```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
data(volcano) # Load volcano contour data
volcano[1:10, 1:10] # Examine volcano dataset (first 10 rows and columns)
volcano3d <- melt(volcano) # Use reshape2 package to melt the data into tidy form
head(volcano3d) # Examine volcano3d dataset (head)
names(volcano3d) <- c("xvar", "yvar", "zvar") # Rename volcano3d columns

ggplot(data = volcano3d, aes(x = xvar, y = yvar, z = zvar)) + geom_contour() # ggplot2
contourplot(zvar~xvar + yvar, data = volcano3d) # lattice
```

# lattice v. ggplot2: tile/image/level plots
```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
ggplot(data = volcano3d, aes(x = xvar, y = yvar, z = zvar)) + geom_tile(aes(fill = zvar)) # ggplot2
levelplot(zvar~xvar + yvar, data = volcano3d) # lattice
```

# lattice: 3D plots
```{r, fig.cap=" ", warning=FALSE}
# Create a subset of the dataset containing only data for France
cloud(Gold~Bronze*Silver, data = northern_hem)

cloud(Gold~Bronze*Silver|Country, data = northern_hem)
```

# lattice v. ggplot2: options [labeling]

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=data, aes(x=year, y=outlays))  +  geom_point() + 
xlab(label="Voter Turnout (%)") + ylab(label="Government Outlays") + 
ggtitle(label="Cool Graph") # ggplot2
xyplot(outlays~year, data=data, xlab="Year", ylab="Government Outlays", main
 ="Cool Graph") # lattice
```

# lattice v. ggplot2: options [axis + size scaling]

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=data, aes(x=year, y=outlays)) + geom_point() # ggplot2
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(size=3) # ggplot2
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(size=1) # ggplot2
xyplot(outlays~year, data=data) # lattice
xyplot(outlays~year, data=data, cex=2) # lattice
xyplot(outlays~year, data=data, cex=.5) # lattice
```

# lattice v. ggplot2: options [graphical parameters]

* Colors
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(color=colors()[145]) # ggplot2
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(color="red") # ggplot2
xyplot(outlays~year, data=data, col=colors()[145]) #lattice
xyplot(outlays~year, data=data, col="red") #lattice
```

* Point Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(shape=3) # ggplot2
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(shape=15) # ggplot2
xyplot(outlays~year, data=data, pch=3) # lattice
xyplot(outlays~year, data=data, pch=15) # lattice
```

* Point Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(shape=3) # ggplot2
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(shape=15) # ggplot2
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(shape="w") # ggplot2
ggplot(data=data, aes(x=year, y=outlays)) + geom_point(shape="$", size=5) # ggplot2
xyplot(outlays~year, data=data, pch=3) # lattice
xyplot(outlays~year, data=data, pch=15) # lattice
xyplot(outlays~year, data=data, pch="w") # lattice
xyplot(outlays~year, data=data, pch="$", cex=2) # lattice
```

* Line Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays)) + 
geom_line(linetype=1) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays)) + 
geom_line(linetype=2) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays)) + 
geom_line(linetype=3) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays)) + 
geom_line(linetype=3, size=1) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays)) + 
geom_line(linetype=3, size=1.5) # ggplot2
ggplot(data=data[data$country=="USA",], aes(x=year, y=outlays)) + 
geom_line(linetype=3, size=2) # ggplot2
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=1) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=2) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3, lwd=2) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3, lwd=3) # lattice
xyplot(outlays~year, data=data[data$country=="USA",], type="l", lty=3, lwd=4) # lattice
```


```{r, fig.cap=" ", warning=FALSE}
# Differences in kind using line types
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_line(aes(linetype=country))
# Differences in kind using point shapes
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point(aes(shape=country))
# Differences in degree using color
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point(aes(color=realgdpgr))
# Differences in degree using point size
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point(aes(size=realgdpgr))
# Multiple non-cartesian aesthetics (differences in kind using color, degree using point size)
ggplot(data=pigs.data, aes(x=year, y=outlays)) + 
  geom_point(aes(color=country,size=realgdpgr))
```

# Fitted lines and curves with ggplot2
```{r, fig.cap=" ", warning=FALSE, message=FALSE}
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point()
# Add linear model (lm) smoother
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point() + 
  geom_smooth(method="lm")
# Add local linear model (loess) smoother, span of 0.75 
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point() + 
  geom_smooth(method="loess", span=.75)
# Add local linear model (loess) smoother, span of 0.25 
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point() + 
  geom_smooth(method="loess", span=.25)
# Add linear model (lm) smoother, no standard error shading 
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point() + 
  geom_smooth(method="lm", se=F)
# Add local linear model (loess) smoother, no standard error shading 
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point() + 
  geom_smooth(method="loess", se=F)
# Add a local linear (loess) smoother for each country
ggplot(data=pigs.data, aes(x=year, y=outlays)) + geom_point(aes(color=country)) + 
  geom_smooth(aes(color=country))
# Add a local linear (loess) smoother for each country, no standard error shading
ggplot(data=pigs.data, aes(x=year, y=outlays)) +
  geom_point(aes(color=country, size=realgdpgr)) + 
  geom_smooth(aes(color=country), se=F)
```

# lattice v. ggplot2: tables

* Both lattice and ggplot2 graphs can be combined using the *grid.arrange()* function in the **gridExtra** package 
```{r, warning=FALSE, fig.cap=" "}
# Initialize gridExtra library
library(gridExtra)
# Create 3 plots to combine in a table
plot1 <- ggplot(data=pigs.data, aes(x=year, y=outlays, color=)) + 
  geom_line(aes(color=country))
plot2 <- ggplot(data=pigs.data, aes(x=year, y=outlays, linetype=)) + 
  geom_line(aes(linetype=country))
plot3 <- ggplot(data=pigs.data, aes(x=year, y=outlays, shape=)) + 
  geom_point(aes(shape=country))
# Call grid.arrange
grid.arrange(plot1, plot2, plot3, nrow=3, ncol=1)
```

# Exporting

Two basic image types

1) **Raster/Bitmap** (.png, .jpeg)

Every pixel of a plot contains its own separate coding; not so great if you want to resize the image

```{r, eval=FALSE}
jpeg(filename="example.png", width=, height=)
plot(x,y)
dev.off()
```

2) **Vector** (.pdf, .ps)

Every element of a plot is encoded with a function that gives its coding conditional on several factors; great for resizing

```{r, eval=FALSE}
pdf(filename="example.pdf", width=, height=)
plot(x,y)
dev.off()
```

# Exporting with lattice v. ggplot

```{r, eval=FALSE}
# Assume we saved our plot is an object called example.plot

# lattice
trellis.device(device="pdf", filename="example.pdf")
print(example.plot)
dev.off()

# ggplot2
ggsave(filename="example.pdf", plot=example.plot, scale=, width=, height=) # ggplot2
```

# Breakout

You're welcome to try out either lattice or ggplot for these questions, but in the solutions we'll focus on the ggplot approach.

For some of these you may want to use a smaller version of the dataset, such as a random subset, `subset <- air[sample(1:nrow(air), 10000, replace = FALSE), ]`.

### Basics

1) Plot a histogram of the flight delays with negative delays set to zero, censoring delay times at a maximum of 60 minutes.

2) Plot the arrival delay against the departure delay as a scatterplot.

3) Clean up your scatterplot with a title and axis labels. Output it as a PDF and see if you'd be comfortable with including it in a report/paper.

4) Make a boxplot of the departure delay as a function of the day of week.

### Using the ideas

5) Create a trellis plot of departure delay boxplots, one per destination for this subset of destinations, `DestSubset <- c('LAX','SEA','PHX','DEN','MSP','JFK','ATL','DFW','IAH', 'ORD')`. Use a 2x5 layout of panels in the plot.

6) Subset the data to flights going to Chicago (ORD) and Houston (IAH). Plot arrival delay against scheduled departure time (CRSDepTime). Now plot so that flights to Chicago are in one color and  those to Houston in another. Use `scale_x_continuous()` and `scale_y_continuous()` to set the x-axis limits to be in the range from 6 am to midnight and the y-axis limits to be in the range (-10, 120).


### Advanced 

7) Create a trellis plot where, for a given destination (see the subset in question 5), each panel uses a) hollow circles to plot departure delay as a function of time of day, and b) a red loess smoother without standard errors to plot the trend in departure delay over time of day.  Limit the time of day shown to 6 am to midnight, and turn off the grey background. Figure out how to use partially-transparent points to reduce the effect of the overplotting of points.


